---
title: "Two years of transducers at HSE"
subtitle: "I dedicate this talk to Nick Howell"
author: "George Moroz"
institute: "Linguistic Convergence Laboratory, HSE University, Moscow"
date: "18 October 2022"
output: 
  beamer_presentation:
    df_print: kable
    latex_engine: xelatex
    citation_package: natbib
    keep_tex: false
    includes:
      in_header: "config/presento.sty"
always_allow_html: true
bibliography: bibliography.bib
urlcolor: colorblue
citecolor: colorblue
csl: "config/apa.csl"
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dev='cairo_pdf')
library(tidyverse)
theme_set(theme_minimal()+theme(legend.position = "bottom", text = element_text(size = 18)))
```

# Tranducers for morphological parsing at HSE

## Tranducers for morphological parsing at HSE

- Nick Howell gave a talk at School of Linguistics' seminar
- In 2020 we started four projects on:
    - [Abaza](https://github.com/dasharakelova/abaza_analyzer) (Daria Arakelova)
    - [Agul](https://github.com/climoy/field_morph) (Roman Klimov)
    - [Bagvalal](https://github.com/ruthenian8/bagvalal/) (Daniil Ignatiev)
    - [Botlikh](https://github.com/ddemain/botlikh-mp-2020/) (Artyom Sinelshikov) \pause
- In summer of 2020 Danya continued his work in Google Summer of Code \pause 
- In 2021 I started five projects on my own ([here](https://agricolamz.github.io/2021.12.02_intro_to_linux_lexd_github_makefiles/) some lectures):
    - [Chamalal](https://github.com/ZinaBudilova/Chamalal_parser) (Zina Budilova)
    - [Botlikh](https://github.com/fkaimphilat/botlikh-morph) (Igor Philatov)
    - [Andi](https://github.com/vbunt/andi) (Lera Buntyakova)
    - [Agul](https://github.com/nstsi/agul) (Nastya Burakova)
    - [Rutul](https://github.com/Hanna-Cupery/rutul) (Hanna Cupery)
    - continuation of [Bagvalal](https://github.com/ruthenian8/bagvalal/) (Daniil Ignatiev)
    - [Even](https://github.com/tbkazakova/even_corpus) (Tanya Kazakova under supervision of Oleg Serikov) \pause
- Result: [@arakelova21; @budilova22; @buntyakova22; @burakova22; @kazakova22;@cupery22]
    
# What is transducer?

## What is transducer?

Trasducers are finite-state networks with two memory tapes of memory that can perform morphological analysis and related tasks. However, it easier to understand via the examples.

- Here is an example of finite-state automaton with one memory tape. It checks whether input string is a word *song*:

```{r}
library(DiagrammeR)
grViz('
  digraph G { rankdir="LR"
  node [fontname="Brill",shape=circle,fontsize=14,fixedsize=true]
  edge [fontname="Brill",fontsize=16]
  0 [label="0"];
  1 [label="1"];
  2 [label="2"];
  3 [label="3"];
  4 [label="4: ✔️",shape=doublecircle];
  5 [label="5: ✘",shape=doublecircle];
  0 -> 1 [label="s"];
  0 -> 5 [label="not s"];
  1 -> 2 [label="o"];
  1 -> 5 [label="not o"];
  2 -> 3 [label="n"];
  2 -> 5 [label="not n"];
  3 -> 4 [label="g"];
  3 -> 5 [label="not g"];
}')
```

## What is transducer?

Trasducers are finite-state networks with two memory tapes of memory that can perform morphological analysis and related tasks. However, it easier to understand via the examples.

- Here is an example of finite-state automaton with one memory tape. It checks whether input string is a word *song*.
- Usually these "not X" arrow are not written:

```{r}
grViz('
  digraph G { rankdir="LR"
  node [fontname="Brill",shape=circle,fontsize=14,fixedsize=true]
  edge [fontname="Brill",fontsize=16]
  0 [label="0"];
  1 [label="1"];
  2 [label="2"];
  3 [label="3"];
  4 [label="4",shape=doublecircle];
  0 -> 1 [label="s"];
  1 -> 2 [label="o"];
  2 -> 3 [label="n"];
  3 -> 4 [label="g"];
}')
```

## What is transducer?

Trasducers are finite-state networks with two memory tapes of memory that can perform morphological analysis and related tasks. However, it easier to understand via the examples.

- Here is an example of finite-state automaton with one memory tape. It checks whether input string is a word *song*.
- Usually this "not X" arrow is not written
- It is possible to code multiple words (*son*, *sons*, *song*):

```{r}
grViz('
  digraph G { rankdir="LR"
  node [fontname="Brill",shape=circle,fontsize=14,fixedsize=true]
  edge [fontname="Brill",fontsize=16]
  0 [label="0"];
  1 [label="1"];
  2 [label="2"];
  3 [label="3",shape=doublecircle];
  4 [label="4",shape=doublecircle];
  5 [label="5",shape=doublecircle];
  0 -> 1 [label="s"];
  1 -> 2 [label="o"];
  2 -> 3 [label="n"];
  3 -> 4 [label="g"];
  3 -> 5 [label="s"];
}')
```


## What is transducer?

Transducers has two memory tapes that can be treated as overwriting:

- `son` becomes `son<n><sg>`
- `song` becomes `song<n><sg>`
- `sons` becomes `son<n><pl>`
- everything else returns an error

```{r}
grViz('
  digraph G { rankdir="LR"
  node [fontname="Brill",shape=circle,fontsize=14,fixedsize=true]
  edge [fontname="Brill",fontsize=16]
  0 [label="0"];
  1 [label="1"];
  2 [label="2"];
  3 [label="3"];
  4 [label="4",shape=doublecircle];
  5 [label="5",shape=doublecircle];
  6 [label="6",shape=doublecircle];
  0 -> 1 [label="s:s"];
  1 -> 2 [label="o:o"];
  2 -> 3 [label="n:n"];
  2 -> 6 [label="n:n<n><sg>"];
  3 -> 4 [label="g:g<n><sg>"];
  3 -> 5 [label="s:<n><pl>"];
}')
```

## Why use transducers?

- they are reversible, so analysis (`sons` → `son<n><pl>`) and generation (`son<n><pl>` → `sons`) can be done with the same transducer \pause
- they can be optimized for the fast search  \pause
- they can be easily combined with other transducers (e. g. transliteration or even translation)

# How to use transducers?

## How to use transducers?

- read [@beesley03; @karttunen92]
- `lexd` --- a finite-state lexicon compiler [@swanson21]
- `twol` --- a tool for (mor)phonology

## `lexd` example (Zilo Andi numerals)

```
PATTERNS
Numerals NumearalMarker

LEXICON Numerals
иᵸшду                 # пять; five
ойлIи                 # шесть; six
гьокьу                # семь; seven
бейкьи                # восемь; eight
гьочIо                # девять; nine

LEXICON NumearalMarker
<num>:гу
```
\pause
```
иᵸшдугу:иᵸшду<num>
ойлIигу:ойлIи<num>
гьокьугу:гьокьу<num>
бейкьигу:бейкьи<num>
гьочIогу:гьочIо<num>
```

## Pipeline for developing morphological transducer

- describe morphology and (mor)phonology using available sources \pause
- compile lexicon with inflectional type annotation \pause
- compile test forms and their analysis based on available data (optional) \pause
    - `чIе<NUM><num><obl.m><epent.m><an.sg><aff> чIегушубо` (Zilo Andi)
- test your transducer against some annotated (or not annotated) corpus (optional)

## Main problems during morphological transducer development

- not usual development environment (at least for our students/not computer linguists)
    - but Tanya Kazakova make it into Google Colab \pause
- time \pause
- lack of resources \pause
- traditions of Apertium people
    - there is a straightforward, but not the shortest way
        - from `чIе<NUM><num><obl.m><epent.m><an.sg><aff> чIегушубо` (Zilo Andi)
        - to two-ɴᴜᴍ-ᴏʙʟ.ᴍ-<ᴀɴ.sɢ>ᴀꜰꜰ \pause
- difference in linguistic descriptions \pause
- difference in languages

# Zilo: from 0 to 40866 wordforms

## Zilo: from 0 to 40866 wordforms

This September I have been in Zilo (Andi) for about 5 days:

- collected a dictionary with inflectional type annotation (more than 700 words)
- created a transducer that analyze/generate 40866 wordforms
    - about 400 nouns, 70 adjectives, 50 adverbs, numerals
    - just nominal and adjective inflection
    - future work: pronouns, verb inflection

# References {.allowframebreaks}
